Evaluating the Decency and Consistency of Data Validation Tests Generated by LLMs

### 2024-03-30 paper/paper.qmd
Update paper and README

---
title: "Evaluating the Decency and Consistency of Data Validation Tests Generated by LLMs"
subtitle: "An application to Canadian political donations data"
author: 
    - name: Rohan Alexander^[Information and Statistical Sciences, University of Toronto. rohan.alexander@utoronto.ca.]
    - name: Lindsay Katz^[The Investigative Journalism Foundation]
    - name: Callandra Moore^[The Investigative Journalism Foundation]
    - name: Michaela Drouillard^[The Investigative Journalism Foundation]
    - name: Michael Wing-Cheung Wong^[Massachusetts Institute of Technology]
    - name: Zane Schwartz^[The Investigative Journalism Foundation]
date: today
date-format: long
abstract: "We investigated the potential of large language models (LLMs) in developing dataset validation tests. We carried out an experiment with 96 conditions each for both GPT-3.5 and GPT-4, examining different prompt scenarios, learning modes, temperature settings, and roles. The prompt scenarios were: 1) Asking for expectations, 2) Asking for expectations with a given context, 3) Asking for expectations after requesting a data simulation, and 4) Asking for expectations with a provided data sample. For learning modes, we tested: 1) zero-shot, 2) one-shot, and 3) few-shot learning. We also tested four temperature settings: 0, 0.4, 0.6, and 1. Furthermore, two distinct roles were considered: 1) helpful assistant, 2) expert data scientist. To gauge consistency, every setup was tested five times. The LLM-generated responses were benchmarked against a gold standard suite, created by an experienced data scientist knowledgeable about the data in question. We find there are considerable returns to the use of few-shot learning, and that the more explicit the data setting can be the better. The best LLM configurations complement, rather than substitute, the gold standard results. This study underscores the value LLMs can bring to the data cleaning and preparation stages of the data science workflow."
thanks: "We thank Monica Alexander, Chris Maddison, Dianne Cook, Yadong Liu, and Will Jettinghoff for helpful suggestions. Code and data: https://github.com/RohanAlexander/evaluating_decency_and_consistency. Contact: rohan.alexander@utoronto.ca. Contributions: RA had the initial idea. RA and CM developed the experimental set-up. ZS established the political donations datasets and obtained funding. LK developed the initial suite of dataset validation tests that we compare the LLMs against. RA wrote the code to interact with the LLMs and did the modelling. RA, MD, and MW evaluated the LLM output. All authors contributed to writing the initial draft as well as improving and finalizing the paper."
bibliography: "bibliography.bib"
format:
  pdf:
    keep-tex: false
header-includes:
  - \usepackage{listings}
  - \lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
number-sections: true
---

@@

The Investigative Journalism Foundation (IJF) created and maintains a dataset related to political donations in Canada. As of September 2023, the dataset comprised 9,204,112 observations and 15 variables. Every day new observations are added, based on newly released donations records made available by provincial and federal elections agencies. This release cycle is variable, with periods of inactivity followed by bursts of multiple releases. @katzandmoore manually construct an extensive suite of automated tests for this dataset. These impose certain minimum standards on the dataset, including: that constituent aspects add to match any total, class is appropriate, and null values are where expected. This suite allows researchers to use the dataset with confidence and ensures that new additions are fit for purpose.

@@

Human evaluators judge the responses produced by the LLMs, rating their decency (1-5, where 1 is the worst and 5 is the best) and their consistency (1-5, where 1 means they are very different and 5 means they are essentially identical). We then build an ordinal regression model with `rstanarm` to explore the relationships that decency and consistency have with prompts, roles, learning modes, and temperature settings.

@@

Our results demonstrate one use for LLMs in a data analysis workflow, outside of just analysis. In particular, data scientists are often concerned that results may be an artifact of some errors in the data cleaning and preparation pipeline. Data validation tests help assuage these concerns but they can be time consuming to produce. This paper shows it is possible to use LLMs to establish an initial suite of tests, which may encourage their use.

The remainder of this paper is structured as follows: @sec-background provides a brief overview of the underlying dataset about which we are writing tests, as well as LLMs. @sec-data details the human coded LLM responses, which is the analysis dataset used in this paper. @sec-model specifies the analysis model used and @sec-results details the estimates. Finally @sec-discussion discusses some of the implications of this study and details a few weaknesses.

@@

The Investigative Journalism Foundation (IJF) is a nonprofit news media outlet that is centered around public interest journalism. Their mandate is to help rebuild trust in Canadian democracy and hold the powerful accountable through data-driven investigative reporting. A core component of the IJF's work is building and actively maintaining eight publicly available databases with information on political donations, registered charities, and political lobbying in Canada. While the information that these databases are built upon are ostensibly public, they are not maintained or available in a way that is widely accessible or conducive to analysis. Further, the format and accessibility of these data vary over time and across jurisdictions, making it difficult to look at temporal or regional differences in the data. In turn, the IJFs collation of these databases and high-quality journalism informed by these data serves as a crucial contribution to rebuilding trust and transparency in Canadian democracy.

@@

While the IJF's database is available in a clean, user-friendly format, the original records upon which it was created were not all accessible in this way. The format of donation records varies across jurisdiction and time. While some are available in readily downloadable spreadsheets, others are available as PDF or HTML files --- the former necessitating the use of optical character recognition (OCR) [@ijfmethods]. To prepare their database for publication, the IJF team performed significant manual cleaning. The majority of this work resulted from the conversion of PDF donation records to rectangular CSV format using OCR. This is prone to scanning-related errors, such as the number 0 being scanned as the letter O. The IJF manually corrected these errors wherever they were identified by comparing the machine-legible OCR output to the original PDF donations record [@ijfmethods].

@@

Despite the cleaning performed by the IJF team, the magnitude of these data coupled with their self-reported nature makes them prone to both human and computational error arising from the parsing process. With over 9.2 million rows in this database, it would be a massive undertaking for the IJF to manually identify all errors and inconsistencies present in their data --- whether that is as minor as an incorrectly formatted date, or as major as a reported donation amount thousands of dollars above the legal limit. To address this challenge, the IJF uses computational tools to assess data quality [@katzandmoore]. The team has developed a suite of comprehensive data tests for all eight databases using Python's `Great Expectations` library. This contains a number of pre-defined functions to test that particular characteristics expected of a dataset hold true in the data at large.

As described by @katzandmoore, the process of developing this test suite was iterative in nature and necessitated significant domain knowledge to develop accurate and valuable tests. For the political donations database, @katzandmoore developed a comprehensive suite of tests pertaining to missingness, formatting, and value expectations in the data. For instance, they set the expectation that for donations made in British Columbia, Ontario, or federally, the `donation_date` entry should not be missing, because those are the only three jurisdictions which collect that variable [@katzandmoore]. They also test that donations data from 2022 align with the 2022 legal donation limits for each jurisdiction, and that, where applicable, the sum of reported monetary and non-monetary contribution amounts add up to the total reported donation amount. @katzandmoore provide complete details on the development and implementation of these tests.

@@

Historically, natural language processing (NLP) models operated in much smaller contexts than today. The first NLP systems were rule-based, relying on hand-crafted linguistic rules based on experts' understanding of languages, and were trained on much smaller datasets.

By the 1990s, advancement in computational capabilities and an increased availability in larger textual datasets allowed for a paradigm shift from rule-based systems to statistical methods. Models learned patterns from data rather than relying on manual rules, and probabilistic models such as Hidden Markov Models (HMMs) and Bayesian networks underpinned new NLP applications. Where rule-based models were brittle and complex, with existing functionalities breaking with the addition of new rules, probabilistic models inferred linguistic patterns from large datasets, which made them much better at scaling and generalizing to solve unseen linguistic challenges.

The transition from statistical methods and traditional machine learning methods to \"deep learning\" came when @bengio2003 introduced neural-network based language models. These models predicted the next word in a sentence, given the previous words. The model of @bengio2003 relied on a representation for words, where every word from the vocabulary is linked with a continuous vector. During training, models refined these vector representations. Words with similar meanings had vectors that were closer together in vector space, which enables the model to understand semantic relationships between words.

Once neural networks were introduced, many significant advancements such as word embeddings [@pennington2014; @mikolov2013], long short-term memory [@hochreiter1997] and Convolutional Neural Networks (CNNs) [@kim2014] followed. The concept of pre-training models and then fine-tuning them for specific tasks (as seen with models like ULMFiT [@howard2018] and ELMo [@peters2018]) started to take root in the late 2010s, which set the stage for transformer-based architectures.

@vaswani2017 introduced transformer architectures, which enabled models to understand complex linguistic patterns and generate text by processing vast amounts of text data. These models, which are characterized by massive parameter counts (among billions or trillions) are trained on diverse corpora, covering cultural nuances and massive varieties of human knowledge. Operating with this broader context has led to LLMs in recent years setting new benchmarks across various NLP tasks, with a potential to reshape many industries.

Large language models (LLMs) are customized and refined through in-context learning using prompting [@ouyang2022]. A prompt is a set of natural language instructions which define the parameters and context for the desired output and may include one or more input-output examples. By using this approach, LLMs can apply their existing knowledge gained from training on various datasets to adapt to new contexts specified by the prompt. The outcomes generated by LLMs are sensitive to the specific phrasing and structure of these natural language instructions. Consequently, there is ongoing work to establish effective prompt patterns [@white2023].

@@

We are interested in the extent to which the LLMs can develop a suite of data validation tests that is similar to a suite developed by an experienced expert data scientist who is familiar with the dataset. To test this, we establish and run a series of experiments where we consider different specifications and then compare the LLM output. In particular, the variables that we consider are summarized here, and the full details are included in [Appendix -@sec-details].

- Four prompts:
    - List the variables of interest -  "index", "amount", "donor_location", "donation_date", "donor_full_name", "donor_type", "political_entity", "political_party", "recipient", "region", "donation_year", "amount_monetary", "amount_non_monetary", "electoral_event", "electoral_district", "added" - and ask for a suite of expectations using the Python package `great_expectations`.
    - As above, plus provide details about each variable. For instance, "amount" is a monetary value that cannot be less than $0. An example observation is "195.46". It is, possible, but unlikely to be more than $1,000.00. It cannot be NA. It should be a numeric. The maximum donation "amount" depends on the value of "region" and "year". For "Federal" is 1675, for "Quebec" is 100 since 2013 and 500 for earlier years, for "British Columbia" is 1309.09, for "Ontario" is 3325, and for "Alberta" is 4300. There is no limit for "Saskatchewan".
    - As above, plus ask the LLM to simulate an example dataset of 1000 observations.
    - Provide the details of the dataset in the first option, as well as an extract of the dataset.
- Three learning modes: zero-, one-, and few-shot learning. In the first, we do not provide any examples; in the second, we provide one example; and in the third, we provide three examples.
- Four temperature values: 0, 0.4, 0.6, and 1.
- Two roles:
    - You are a helpful assistant.
    - You are a highly-trained, experienced, data scientist who is an expert at writing readable, correct, Python code.

This combination of variables and options results in 96 different prompt situations. We run these through both GPT-3.5 and GPT-4, using the API. For every combination we ask for five responses to understand variation.

@@

To bring the three evaluators together we consider the median response. This then allows us to examine whether there is a relationship between decency and consistency (@fig-decencyvsconsistency). (Details about coder-specific responses are provided in [Appendix -@sec-intercoder].)

```{r}
#| echo: false
#| label: fig-decencyvsconsistency
#| fig-cap: "How decency and consistency compare with each other"
#| warning: false

data |>
  ggplot(aes(x = decency_median, y = consistency_median)) +
  geom_jitter(width = 0.15,
              height = 0.15,
              alpha = 0.5) +
  theme_minimal() +
  labs(x = "Decency", y = "Consistency") +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

@fig-version examines how decency and consistency differ based on whether GPT-3.5 or GPT-4 is used. Unexpectedly, GPT-4 has fewer responses rated 5/5 for decency, compared with GPT-3.5 (@fig-version-1). GPT-3.5 has fewer responses rated 2/5, but overall the mean decency response for GPT-3.5 is 3.26, while the mean decency of the responses generated by GPT-4 is 3.19. The consistency is not too different between the two versions, with GPT-3.5 having an average of 3.21, while GPT-4 has an average of 3.33. GPT-4 appears to have fewer responses that are completely identical (@fig-version-2).

@@

```{r}
#| echo: false
#| label: fig-version
#| fig-cap: "How decency and consistency change based on whether GPT-3.5 or GPT-4 is used"
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

data |> 
    ggplot(aes(x = decency_median, fill = Version)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Decency", y = "Number of occurrences"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")

data |> 
    ggplot(aes(x = consistency_median, fill = Version)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Consistency", y = "Number of occurrences"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")
```

@@

There appears to be considerable difference in terms of how the prompts are associated with decency. In particular, "Name" is never associated with a 5/5 rating (@fig-prompt-1). Surprisingly, however, the most informative prompt, "Example", is also never associated with a 5/5 rating. Instead it is "Describe" and "Simulate" that tend to be associated with better decency ratings. This is reflected in the averages, which are 3.15, 3.31, 3.27, and 3.17 for "Name", "Describe", "Simulate", and "Example", respectively.

The pattern is not as clear when it comes to consistency (@fig-prompt-2). All four have similar averages, at 3.42, 3.10, 3.21, and 3.35 for "Name", "Describe", "Simulate", and "Example", respectively. That said, it is clear that a wider variety of responses (as denoted by lower consistency ratings), are rarely seen for "Name" and "Describe".

@@

```{r}
#| echo: false
#| label: fig-prompt
#| fig-cap: "How decency and consistency change based on the type of prompt used"
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

data |> 
    ggplot(aes(x = decency_median, fill = Prompt_n)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Decency", y = "Number of occurrences", fill = "Prompt type"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")

data |> 
    ggplot(aes(x = consistency_median, fill = Prompt_n)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Consistency", y = "Number of occurrences", fill = "Prompt type"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")
```

Temperature is a parameter that varies from 0 to 1, that we can use to manipulate how random the LLM is. At high temperatures, the LLM will produce a wider variety of responses. At lower temperatures it will focus on the single most likely response. Higher temperature should be associated with a wider variety of LLM responses.

@@

In contrast, as expected temperature has an effect on consistency. Temperature values of 0 are associated with ratings of high consistency, and higher temperatures are associated with lower consistency (@fig-temperature-2). Their means differ considerably, with 4.73, 3.40, 2.69, and 2.27 for temperature values of 0, 0.4, 0.6, and 1, respectively.

@@

```{r}
#| echo: false
#| label: fig-temperature
#| fig-cap: "How decency and consistency change based on temperature"
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

data |> 
    ggplot(aes(x = decency_median, fill = Temperature)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Decency", y = "Number of occurrences"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")

data |> 
    ggplot(aes(x = consistency_median, fill = Temperature)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Consistency", y = "Number of occurrences"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")
```

Role is an aspect of a prompt that is provided to the LLM before the main prompt content. We considered two different roles, one that positioned the LLM as a helpful assistant, and the other that positioned the LLM as an experienced data scientist (@fig-role). We were expecting that the expert role would result in better code, but there was no obvious difference in terms of decency (@fig-role-1) or consistency (@fig-role-2). Their mean decency did not differ by much in either case with 3.12 and 3.32 for 'Helpful' and 'Expert', respectively, while mean consistency was 3.35 and 3.19.

@@

```{r}
#| echo: false
#| label: fig-role
#| fig-cap: "How decency and consistency change based on whether role is 'Helpful' or 'Expert'"
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

data |> 
    ggplot(aes(x = decency_median, fill = Role_n)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Decency", y = "Number of occurrences", fill = "Role"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")

data |> 
    ggplot(aes(x = consistency_median, fill = Role_n)) +
    geom_histogram(position = "dodge2", binwidth = 1) +
    theme_minimal() +
    labs(
        x = "Consistency", y = "Number of occurrences", fill = "Role"
    ) +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "bottom")
```

>>>>

---
title: "Evaluating the Decency and Consistency of Data Validation Tests Generated by LLMs"
subtitle: "An application to Canadian political donations data"
author: 
    - name: Rohan Alexander^[Information and Statistical Sciences, University of Toronto. rohan.alexander@utoronto.ca.]
    - name: Lindsay Katz^[The Investigative Journalism Foundation]
    - name: Callandra Moore^[The Investigative Journalism Foundation]
    - name: Michaela Drouillard^[The Investigative Journalism Foundation]
    - name: Michael Wing-Cheung Wong^[Massachusetts Institute of Technology]
    - name: Zane Schwartz^[The Investigative Journalism Foundation]
date: today
date-format: long
abstract: "We investigated whether large language models (LLMs) can develop data validation tests. We considered 96 conditions each for both GPT-3.5 and GPT-4, examining different prompt scenarios, learning modes, temperature settings, and roles. The prompt scenarios were: 1) Asking for expectations, 2) Asking for expectations with a given context, 3) Asking for expectations after requesting a data simulation, and 4) Asking for expectations with a provided data sample. The learning modes were: 1) zero-shot, 2) one-shot, and 3) few-shot learning. We also tested four temperature settings: 0, 0.4, 0.6, and 1. And the two distinct roles were: 1) helpful assistant, 2) expert data scientist. To gauge consistency, every setup was tested five times. The LLM-generated responses were benchmarked against a gold standard data validation suite, created by an experienced data scientist knowledgeable about the data in question. We find there are considerable returns to the use of few-shot learning, and that the more explicit the data setting can be the better, to a point. The best LLM configurations complement, rather than substitute, the gold standard results. This study underscores the value LLMs can bring to the data cleaning and preparation stages of the data science workflow, but highlights that they need considerable evaluation by experienced analysts."
thanks: "We thank Monica Alexander, Chris Maddison, Dianne Cook, Yadong Liu, and Will Jettinghoff for helpful suggestions. Code and data: https://github.com/RohanAlexander/evaluating_decency_and_consistency. Contact: rohan.alexander@utoronto.ca. Contributions: RA had the initial idea. RA and CM developed the experimental set-up. ZS established the political donations datasets and obtained funding. LK developed the initial suite of dataset validation tests that we compare the LLMs against. RA wrote the code to interact with the LLMs and did the modelling. RA, MD, and MW evaluated the LLM output. All authors contributed to writing the initial draft as well as improving and finalizing the paper."
bibliography: "bibliography.bib"
format:
  pdf:
    keep-tex: false
header-includes:
  - \usepackage{listings}
  - \lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
number-sections: true
---

@@

The Investigative Journalism Foundation (IJF) created and maintains a dataset related to political donations in Canada. As of September 2023, the dataset comprised 9,204,112 observations and 15 variables. Every day new observations are added, based on newly released donations records made available by provincial and federal elections agencies. This release cycle is variable, with periods of inactivity followed by bursts of multiple releases. @katzandmoore manually construct an extensive suite of automated tests for this dataset. These impose certain minimum standards on the dataset, including: that constituent aspects add to match any total, class is appropriate, and null values are as expected. This suite allows researchers to use the dataset with confidence and ensures that new additions are fit for purpose.

@@

Three human coders judge the responses produced by the LLMs, rating their decency (1-5, where 1 is the worst and 5 is the best) and their consistency (1-5, where 1 means they are very different and 5 means they are essentially identical). We then build an ordinal regression model with `rstanarm` to explore the relationships that decency and consistency have with prompts, roles, learning modes, and temperature settings.

@@

Our results demonstrate one use for LLMs in production, outside of just analysis. In particular, data scientists are often concerned that results may be an artifact of some errors in the data cleaning and preparation pipeline. Data validation tests help assuage these concerns but they can be time consuming to produce. This paper shows it is possible to use LLMs to establish an initial suite of tests, which may encourage their use. But the variability of responses means these are best seen as a starting place.

The remainder of this paper is structured as follows: @sec-background provides a brief overview of the underlying dataset about which we are writing tests, as well as LLMs. @sec-data details the human coded LLM responses, which is the analysis dataset used in this paper. @sec-model specifies the analysis model used and @sec-results details the estimates and results. Finally, @sec-discussion discusses some of the implications of this study and details a few weaknesses.

@@

The Investigative Journalism Foundation (IJF) is a nonprofit news media outlet that is centered around public interest journalism. Their mandate is to help rebuild trust in Canadian democracy and hold the powerful accountable through data-driven investigative reporting. A core component of the IJF's work is building and actively maintaining nine publicly available databases with information on political donations, registered charities, and political lobbying in Canada. While the information that these databases are built upon are ostensibly public, they are typically not maintained or available in a way that is widely accessible or conducive to analysis. Further, the format and accessibility of these data vary over time and across jurisdictions, making it difficult to look at temporal or regional differences in the data. The IJF's collation of these databases, and the subsequent high-quality journalism informed by these data, serves as a crucial contribution to rebuilding trust and transparency in Canadian democracy.

@@

While the IJF's database is available in a clean, user-friendly format, the original records upon which it was created were not all accessible in this way. The format of donation records varies across jurisdiction and time. While some are available in readily downloadable spreadsheets, others are available as PDF or HTML files---the former necessitating the use of optical character recognition (OCR) [@ijfmethods]. To prepare their database for publication, the IJF team performed significant manual cleaning. The majority of this work resulted from the conversion of PDF donation records to rectangular CSV format using OCR. This is prone to scanning-related errors, such as the number 0 being scanned as the letter O. The IJF manually corrected these errors wherever they were identified by comparing the machine-legible OCR output to the original PDF donations record [@ijfmethods].

@@

Despite the cleaning performed by the IJF team, the magnitude of these data coupled with their self-reported nature makes them prone to both human and computational error arising from the parsing process. With over 9.2 million rows in this database, it would be difficult for the IJF to manually identify all errors and inconsistencies present in their data---whether that is as minor as an incorrectly formatted date, or as major as a reported donation amount thousands of dollars above the legal limit. To address this challenge, the IJF uses computational tools to assess data quality [@katzandmoore]. The team has developed a suite of comprehensive data tests for all nine databases using Python's `Great Expectations` library. This contains a number of pre-defined functions to test that particular characteristics expected of a dataset hold true in the data at large.

As described by @katzandmoore, the process of developing this test suite was iterative and necessitated significant domain knowledge to develop accurate and valuable tests. For the political donations database, @katzandmoore developed a suite of tests pertaining to missingness, formatting, and value expectations in the data. For instance, they set the expectation that for donations made in British Columbia, Ontario, or federally, the `donation_date` entry should not be missing, because those three jurisdictions collect that variable [@katzandmoore]. They also test that, say, donations data from 2022 align with the 2022 legal donation limits for each jurisdiction, and that, where applicable, the sum of reported monetary and non-monetary contribution amounts add up to the total reported donation amount. @katzandmoore provide further details on the development and implementation of these tests.

@@

Historically, natural language processing (NLP) models operated in smaller contexts than today. The first NLP systems were rule-based, relying on hand-crafted linguistic rules based on experts' understanding of languages, and were trained on much smaller datasets.

By the 1990s, advancement in computational capabilities and an increased availability in larger textual datasets allowed for a paradigm shift from rule-based systems to statistical methods. Models learned patterns from data rather than relying on manual rules, and probabilistic models such as Hidden Markov Models (HMMs) and Bayesian networks underpinned new NLP applications. Where rule-based models were brittle and complex, with existing functionalities breaking with the addition of new rules, probabilistic models inferred linguistic patterns from large datasets, which made them better at scaling and generalizing to solve unseen linguistic challenges.

The transition from statistical methods and traditional machine learning methods to "deep learning" came when @bengio2003 introduced neural network based language models. These models predicted the next word in a sentence, given the previous words. The model of @bengio2003 relied on a representation for words, where every word from the vocabulary is linked with a continuous vector. During training, models refined these vector representations. Words with similar meanings had vectors that were closer together in vector space, which enabled the model to understand semantic relationships between words.

Once neural networks were introduced, many significant advancements such as word embeddings [@pennington2014; @mikolov2013], long short-term memory [@hochreiter1997] and Convolutional Neural Networks (CNNs) [@kim2014] followed. The concept of pre-training models and then fine-tuning them for specific tasks, as seen with models like ULMFiT [@howard2018] and ELMo [@peters2018], started to take root in the late 2010s, which set the stage for transformer-based architectures.

@vaswani2017 introduced transformer architectures, which enabled models to understand complex linguistic patterns and generate text by processing vast amounts of text data. These models, which are characterized by parameter counts among billions or trillions, are trained on diverse corpora, covering cultural nuances and varieties of human knowledge. Operating with this broader context has led to LLMs in recent years setting new benchmarks across various NLP tasks, with a potential to reshape many industries.

LLMs are customized and refined through in-context learning using prompting [@ouyang2022]. A prompt is a set of natural language instructions which define the parameters and context for the desired output and may include one or more input-output examples. By using this approach, LLMs can apply their existing knowledge gained from training on various datasets to adapt to new contexts specified by the prompt. The outcomes generated by LLMs are sensitive to the specific phrasing and structure of these natural language instructions. Consequently, there is ongoing work to establish effective prompt patterns [@white2023].

@@

We are interested in the extent to which the LLMs can develop a suite of data validation tests that is similar to a suite developed by an experienced expert data scientist who is familiar with the dataset. To test this, we establish and run a series of experiments where we consider different specifications and then compare the LLM output. The variables that we consider are summarized here, and the full details are included in [Appendix -@sec-details].

- Four prompts:
    - List the variables of interest---"index", "amount", "donor_location", "donation_date", "donor_full_name", "donor_type", "political_entity", "political_party", "recipient", "region", "donation_year", "amount_monetary", "amount_non_monetary", "electoral_event", "electoral_district", "added"---and ask for a suite of expectations using the Python package `great_expectations`.
    - As above, plus provide details about each variable. For instance, "amount" is a monetary value that cannot be less than $0. An example observation is "195.46". It is, possible, but unlikely to be more than $1,000.00. It cannot be NA. It should be a numeric. The maximum donation "amount" depends on the value of "region" and "year". For "Federal" is 1675, for "Quebec" is 100 since 2013 and 500 for earlier years, for "British Columbia" is 1309.09, for "Ontario" is 3325, and for "Alberta" is 4300. There is no limit for "Saskatchewan".
    - As above, plus ask the LLM to simulate an example dataset of 1000 observations.
    - Provide the details of the dataset in the first option, as well as an extract of the dataset.
- Three learning modes: zero-, one-, and few-shot learning. In the first, we do not provide any examples; in the second, we provide one example; and in the third, we provide three examples.
- Four temperature values: 0, 0.4, 0.6, and 1.
- Two roles:
    - You are a helpful assistant.
    - You are a highly-trained, experienced, data scientist who is an expert at writing readable, correct, Python code.

This combination of variables and options results in 96 different prompt situations. We run these through both GPT-3.5 and GPT-4, using the OpenAI API. For every combination we ask for five responses to understand variation.

@@

To bring the three evaluators together we consider the median response. This then allows us to examine whether there is a relationship between decency and consistency (@fig-decencyvsconsistency). There is a broadly positive relationship. Details about coder-specific responses are provided in [Appendix -@sec-intercoder].

```{r}
#| echo: false
#| label: fig-decencyvsconsistency
#| fig-cap: "How decency and consistency compare with each other. Higher decency means the LLM responses correspond more closely with the gold standard. Higher consistency means all five responses were similar to each other."
#| warning: false

data |>
  ggplot(aes(x = decency_median, y = consistency_median)) +
  geom_jitter(width = 0.15,
              height = 0.15,
              alpha = 0.5) +
  geom_smooth(method = lm, se = TRUE, linetype = "dashed", color = "black") +
  theme_minimal() +
  labs(x = "Decency", y = "Consistency") +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

@fig-version illustrates how decency and consistency differ based on whether GPT-3.5 or GPT-4 is used. Unexpectedly, GPT-4 has fewer responses rated 5/5 for decency, compared with GPT-3.5 with 3% compared with 9% (@fig-version-1). Overall, the mean decency response for GPT-3.5 is 3.26, while the mean decency of the responses generated by GPT-4 is 3.19. The consistency is not too different between the two versions, with GPT-3.5 having an average of 3.21, while GPT-4 has an average of 3.33. Unexpectedly, GPT-4 appears to have fewer responses that are essentially identical i.e. 5/5 (@fig-version-2).

@@

```{r}
#| echo: false
#| label: fig-version
#| fig-cap: "How decency and consistency change based on whether GPT-3.5 or GPT-4 is used. For decency, 1 is the worst and 5 is the best. For consistency 5 means all responses were essentially the same, while 1 means at least one was quite different."
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

# data |> 
#     ggplot(aes(x = decency_median, fill = Version)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Decency", y = "Number of occurrences"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")

# data |> 
#     ggplot(aes(x = consistency_median, fill = Version)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Consistency", y = "Number of occurrences"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")
```

@@

There is a difference in terms of how the prompts are associated with decency (@fig-prompt-1). In particular, "Name" is almost never associated with a 5/5 rating, and it has many responses rated 3/5. Surprisingly the most informative prompt, "Example", is also never associated with a 5/5 rating. Instead it is "Describe" and "Simulate" that tend to be associated with better decency ratings. This is reflected in the averages, which are 3.15, 3.31, 3.27, and 3.17 for "Name", "Describe", "Simulate", and "Example", respectively.

The pattern is less clear when it comes to consistency (@fig-prompt-2). All four have similar averages, at 3.42, 3.10, 3.21, and 3.35 for "Name", "Describe", "Simulate", and "Example", respectively. A wider variety of responses (as denoted by lower consistency ratings), is rarely seen for "Name" and "Describe". It is surprising that "Name" should have the highest average.

@@

```{r}
#| echo: false
#| label: fig-prompt
#| fig-cap: "How decency and consistency change based on the type of prompt used. Name means essentially only the variables of interest are listed. Describe means some detail about each variable is provided. Simulate means the LLM is asked to first think about simulating the variables. Example means an extract of the dataset is provided."
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

# data |> 
#     ggplot(aes(x = decency_median, fill = Prompt_n)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Decency", y = "Number of occurrences", fill = "Prompt type"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")

# data |> 
#     ggplot(aes(x = consistency_median, fill = Prompt_n)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Consistency", y = "Number of occurrences", fill = "Prompt type"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")
```

Temperature is a parameter that varies from 0 to 1, that we use to manipulate how random a LLM is. At high temperatures, an LLM will produce a wider variety of responses. At lower temperatures it will focus on the single most likely response. Higher temperature should be associated with a wider variety of LLM responses.

@@

In contrast, a sizeable different in consistency can be seen with different temperatures values (@fig-temperature-2). Temperature values of 0 are associated with ratings of high consistency, and higher temperatures are associated with lower consistency. Their means differ considerably, with 4.73, 3.40, 2.69, and 2.27 for temperature values of 0, 0.4, 0.6, and 1, respectively. The one unexpected aspect is temperature of 0.6, which has an outsized number of responses with decency of 1/5 or 2/5.

@@

```{r}
#| echo: false
#| label: fig-temperature
#| fig-cap: "How decency and consistency change based on temperature. Temperature of 0 means that less variation is expected from the outputs, while temperature of 1 means considerable variation is expected."
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

# data |> 
#     ggplot(aes(x = decency_median, fill = Temperature)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Decency", y = "Number of occurrences"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")

# data |> 
#     ggplot(aes(x = consistency_median, fill = Temperature)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Consistency", y = "Number of occurrences"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")
```

Role is an aspect of a prompt that is provided to the LLM before the main prompt content. We considered two different roles, one that positioned the LLM as a helpful assistant, and the other that positioned the LLM as an experienced data scientist (@fig-role). We were expecting that the expert role would result in better code, but there was no obvious difference in terms of decency (@fig-role-1) or consistency (@fig-role-2). Although it is notable that there are no responses rated 1/5 with the "Expert" role. Their mean decency did not differ by much in either case with 3.12 and 3.32 for 'Helpful' and 'Expert', respectively, while mean consistency was 3.35 and 3.19.

@@

```{r}
#| echo: false
#| label: fig-role
#| fig-cap: "How decency and consistency change based on whether role is 'Helpful' or 'Expert'. This is considered before the main prompt content and is designed to position the LLM as having a certain role."
#| fig-subcap: ["Decency","Consistency"]
#| warning: false
#| layout-ncol: 2

@@

# data |> 
#     ggplot(aes(x = decency_median, fill = Role_n)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Decency", y = "Number of occurrences", fill = "Role"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")

# data |> 
#     ggplot(aes(x = consistency_median, fill = Role_n)) +
#     geom_histogram(position = "dodge2", binwidth = 1) +
#     theme_minimal() +
#     labs(
#         x = "Consistency", y = "Number of occurrences", fill = "Role"
#     ) +
#     scale_fill_brewer(palette = "Set1") +
#     theme(legend.position = "bottom")
```
